---
title: "Exploratory Analysis of Medication Use and Patient Mortality in MIMIC-IV"
date: "2026-02-26"
output: html_document
---

Exploratory Data Analysis

```{r}
library(tidyverse)
library(gt)
library(here)
```

## Outcome: 30-Day Mortality

```{r}
#| fig-cap: "Distribution of 30-day mortality in the analytic cohort"
analysis_df <- read_csv("data/processed-data/analysis_dataset.csv", show_col_types = FALSE)
analysis_df |>
  mutate(Outcome = if_else(mortality_30day == 1, "Died (≤30 days)", "Survived")) |>
  count(Outcome) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = Outcome, y = n, fill = Outcome)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(n, "\n(", scales::percent(pct, 0.1), ")")),
            vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Died (≤30 days)" = "#d62728",
                                "Survived"        = "#2ca02c")) +
  labs(title = "30-Day Mortality Distribution",
       x = NULL, y = "Number of Admissions") +
  theme(plot.title = element_text(face = "bold"))

ggsave("results/figures/day-30-mortality.png", 
       width = 8, height = 5, dpi = 300)
```

30-day mortality is low in this cohort, there are about 1 in 20 admissions that resulted in death within 30 days.

## Demographics

### Age Distribution by Outcome

```{r}
#| fig-cap: "Age distribution by 30-day mortality"

analysis_df |>
  mutate(Outcome = if_else(mortality_30day == 1, "Died", "Survived")) |>
  ggplot(aes(x = age_at_admission, fill = Outcome)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("Died" = "#d62728", "Survived" = "#2ca02c")) +
  labs(title = "Age Distribution by 30-Day Mortality",
       x = "Age at Admission (years)", y = "Count", fill = "Outcome")

ggsave("results/figures/age-distribution-mortality.png", 
       width = 8, height = 5, dpi = 300)
```

The age distribution shows that most people in the study are older adults. 

### Mortality Rate by Age Group

```{r}
#| fig-cap: "30-day mortality rate by age group"

analysis_df |>
  group_by(age_group) |>
  summarise(
    n          = n(),
    deaths     = sum(mortality_30day),
    mort_rate  = deaths / n
  ) |>
  ggplot(aes(x = age_group, y = mort_rate)) +
  geom_col(fill = "#1f77b4", width = 0.6) +
  geom_text(aes(label = scales::percent(mort_rate, 0.1)), vjust = -0.3) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "30-Day Mortality Rate by Age Group",
       x = "Age Group", y = "Mortality Rate")
ggsave("results/figures/mortality_by_age.png", 
       width = 8, height = 5, dpi = 300)
```

30 day mortality increased drastically by as the age group increased.

### Mortality Rate by Gender

```{r}
#| fig-cap: "30-day mortality rate by gender"

tbl_gender <- analysis_df |>
  group_by(gender) |>
  summarise(
    n         = n(),
    deaths    = sum(mortality_30day),
    mort_rate = deaths / n
  )

tbl_gender |>
  gt() |>
  fmt_percent(columns = mort_rate, decimals = 1) |>
  cols_label(
    gender    = "Gender",
    n         = "Admissions",
    deaths    = "Deaths",
    mort_rate = "30-Day Mortality Rate"
  )

saveRDS(tbl_gender, here("results", "tables", "mortality_by_gender.rds"))

```

## Medication Class Usage

### Most Frequently Prescribed Medication Classes

```{r}
#| fig-cap: "Frequency of medication classes across admissions"
med_class_map <- tribble(
  ~pattern,               ~med_class,
  "ASPIRIN",              "Antiplatelet",
  "CLOPIDOGREL",          "Antiplatelet",
  "HEPARIN",              "Anticoagulant",
  "WARFARIN",             "Anticoagulant",
  "ENOXAPARIN",           "Anticoagulant",
  "METOPROLOL",           "Beta-Blocker",
  "CARVEDILOL",           "Beta-Blocker",
  "ATENOLOL",             "Beta-Blocker",
  "LISINOPRIL",           "ACE Inhibitor",
  "ENALAPRIL",            "ACE Inhibitor",
  "CAPTOPRIL",            "ACE Inhibitor",
  "AMLODIPINE",           "Calcium Channel Blocker",
  "DILTIAZEM",            "Calcium Channel Blocker",
  "FUROSEMIDE",           "Diuretic",
  "SPIRONOLACTONE",       "Diuretic",
  "VANCOMYCIN",           "Antibiotic",
  "PIPERACILLIN",         "Antibiotic",
  "CEFAZOLIN",            "Antibiotic",
  "METRONIDAZOLE",        "Antibiotic",
  "CIPROFLOXACIN",        "Antibiotic",
  "INSULIN",              "Insulin",
  "MORPHINE",             "Opioid",
  "FENTANYL",             "Opioid",
  "HYDROMORPHONE",        "Opioid",
  "OXYCODONE",            "Opioid",
  "NOREPINEPHRINE",       "Vasopressor",
  "VASOPRESSIN",          "Vasopressor",
  "DOPAMINE",             "Vasopressor",
  "EPINEPHRINE",          "Vasopressor",
  "PROPOFOL",             "Sedative",
  "MIDAZOLAM",            "Sedative",
  "LORAZEPAM",            "Sedative",
  "DEXAMETHASONE",        "Corticosteroid",
  "PREDNISONE",           "Corticosteroid",
  "METHYLPREDNISOLONE",   "Corticosteroid",
  "ATORVASTATIN",         "Statin",
  "SIMVASTATIN",          "Statin",
  "ROSUVASTATIN",         "Statin"
)

# Get binary med class columns
med_cols <- tolower(unique(med_class_map$med_class)) |>
  str_replace_all("[ -]", "_") |>
  intersect(names(analysis_df))

analysis_df |>
  select(all_of(med_cols)) |>
  summarise(across(everything(), sum)) |>
  pivot_longer(everything(), names_to = "med_class", values_to = "n_admissions") |>
  mutate(
    med_class = str_to_title(str_replace_all(med_class, "_", " ")),
    pct       = n_admissions / nrow(analysis_df)
  ) |>
  arrange(desc(n_admissions)) |>
  ggplot(aes(x = reorder(med_class, n_admissions), y = pct)) +
  geom_col(fill = "#1f77b4") +
  geom_text(aes(label = scales::percent(pct, 0.1)), hjust = -0.1, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format(), expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Medication Class Usage Among Admitted Patients",
       x = NULL, y = "Proportion of Admissions")

ggsave("results/figures/perscriptoin-frequency.png", 
       width = 8, height = 5, dpi = 300)
```

The graph shows that many particpants used cardiovascular and anticoagulations medicine. 

## 4.4 Medication Class vs. 30-Day Mortality

This is the core exploratory analysis: comparing 30-day mortality rates across patients who did vs. did not receive each medication class.

```{r}
#| fig-cap: "30-day mortality rate by medication class (unadjusted)"

med_mortality <- map_dfr(med_cols, function(col) {
  analysis_df |>
    group_by(received = .data[[col]]) |>
    summarise(n = n(), deaths = sum(mortality_30day), .groups = "drop") |>
    mutate(
      med_class  = str_to_title(str_replace_all(col, "_", " ")),
      mort_rate  = deaths / n,
      received   = if_else(received == 1L, "Received", "Not Received")
    )
})

med_mortality |>
  ggplot(aes(x = reorder(med_class, mort_rate), y = mort_rate, fill = received)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Received" = "#d62728", "Not Received" = "#aec7e8")) +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(title = "Unadjusted 30-Day Mortality Rate by Medication Class",
       subtitle = "Patients who received vs. did not receive each class",
       x = NULL, y = "30-Day Mortality Rate", fill = NULL) +
  theme(legend.position = "top")

ggsave("results/figures/unadjusted-perscriptoin-frequency.png", 
       width = 8, height = 5, dpi = 300)
```

### Summary Table

```{r}
med_mortality |>
  filter(received == "Received") |>
  arrange(desc(mort_rate)) |>
  select(med_class, n, deaths, mort_rate) |>
  gt() |>
  fmt_percent(columns = mort_rate, decimals = 1) |>
  fmt_number(columns = c(n, deaths), use_seps = TRUE, decimals = 0) |>
  cols_label(
    med_class = "Medication Class",
    n         = "Admissions Receiving",
    deaths    = "Deaths (30-day)",
    mort_rate = "Mortality Rate"
  ) |>
  tab_header(title = "30-Day Mortality by Medication Class (Unadjusted)")

saveRDS(med_mortality, here("results", "tables", "medication_mortality_summary.rds"))

```

Mortality rates were higher with patients taking vasopressors, sedatives, and diruetics; this could be because the severity of their illness.


## Number of Medication Classes per Admission

Polypharmacy (receiving many medication classes) may itself be associated with disease severity and mortality.

```{r}
#| fig-cap: "Number of medication classes received vs. 30-day mortality"

analysis_df |>
  mutate(n_med_classes = rowSums(across(all_of(med_cols)))) |>
  group_by(n_med_classes) |>
  summarise(
    n         = n(),
    mort_rate = mean(mortality_30day)
  ) |>
  ggplot(aes(x = n_med_classes, y = mort_rate)) +
  geom_point(aes(size = n), color = "#d62728", alpha = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "#1f77b4") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_size_continuous(name = "N Admissions") +
  labs(title = "Polypharmacy and 30-Day Mortality",
       x = "Number of Medication Classes Received",
       y = "30-Day Mortality Rate")

ggsave("results/figures/polypharmacy-frequency.png", 
       width = 8, height = 5, dpi = 300)
```

Mortality increased as the number of medications increased, this also could be explained by the increasing severity of illness as more medications were needed.