---
title: "MIMIC-IV Data Processing Pipeline"
author: "Riley Herber and Rebecca Basta"
date: "2026-02-26"
output: html_document
---

This document covers all data loading, cleaning, and wrangling or the project examining the relationship between medication use and 30-day mortality in the MIMIC-IV database.

**Primary outcome:** 30-day mortality (binary: died within 30 days of admission)  
**Primary predictor:** Medication use (drug name / medication class)  
**Covariates:** Age at admission, gender  


```{r}
library(tidyverse)   # data wrangling & ggplot2
library(lubridate)   # date handling
library(janitor)     # clean_names(), tabyl()
library(skimr)       # skim() for quick summaries
library(gt)          # nice tables
library(here)        # relative file paths

theme_set(theme_minimal(base_size = 13))
```

1. Data Loading

The tables used are:

Table	Description
patients.csv	Demographics (gender, anchor_age, dod)
admissions.csv	Admission/discharge times, race
prescriptions.csv	Medication orders per hospital stay


```{r}
# Optional sampling for speed
set.seed(123)

admissions <- read_csv(here("data", "raw-data", "admissions_syn.csv"), show_col_types = FALSE)
prescriptions <- read_csv(here("data", "raw-data", "prescriptions_syn.csv"), show_col_types = FALSE)
patients <- read_csv(here("data", "raw-data", "patients_syn.csv"), show_col_types = FALSE)


sampled_ids <- admissions |>
  slice_sample(prop = 0.8) |>
  pull(hadm_id)

admissions    <- admissions    |> filter(hadm_id %in% sampled_ids)
prescriptions <- prescriptions |> filter(hadm_id %in% sampled_ids)
patients      <- patients      |> filter(subject_id %in% admissions$subject_id)

patients      <- clean_names(patients)
admissions    <- clean_names(admissions)
prescriptions <- clean_names(prescriptions)

cat("patients:", nrow(patients), "\n")
cat("admissions:", nrow(admissions), "\n")
cat("prescriptions:", nrow(prescriptions), "\n")
```


2. Data Cleaning
2.1 Patients Table

```{r}
patients_clean <- patients |>
  select(subject_id, gender, anchor_age, anchor_year, dod) |>
  mutate(dod = as_date(dod))

skim(patients_clean)

```

2.2 Admissions Table
```{r}
admissions_clean <- admissions |>
  select(subject_id, hadm_id, admittime, dischtime, admission_type, race) |>
  mutate(
    admittime  = ymd_hms(admittime),
    dischtime  = ymd_hms(dischtime),
    los_days   = as.numeric(difftime(dischtime, admittime, units = "days"))
  ) |>
  filter(los_days >= 0, los_days < 365)

cat("Admissions after LOS filter:", nrow(admissions_clean), "\n")

```


3. Compute 30-Day Mortality

```{r}
admissions_mortality <- admissions_clean |>
  left_join(
    patients_clean |> select(subject_id, dod, gender, anchor_age),
    by = "subject_id"
  ) |>
  mutate(
    days_to_death = as.numeric(difftime(dod, admittime, units = "days")),
    mortality_30day = case_when(
      !is.na(days_to_death) & days_to_death >= 0 & days_to_death <= 30 ~ 1L,
      TRUE ~ 0L
    )
  )

admissions_mortality |>
  count(mortality_30day) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  gt()

```


4. Prescriptions Table
```{r}
prescriptions_clean <- prescriptions |>
  select(subject_id, hadm_id, drug, dose_val_rx, dose_unit_rx) |>
  filter(!is.na(drug), drug != "") |>
  mutate(
    drug_std = str_squish(str_to_upper(drug))
  )

cat("Prescriptions after cleaning:", nrow(prescriptions_clean), "\n")
cat("Unique drugs:", n_distinct(prescriptions_clean$drug_std), "\n")

```


5. Assign Medication Classes

```{r}
med_class_map <- tribble(
  ~pattern, ~med_class,
  "ASPIRIN","Antiplatelet",
  "HEPARIN","Anticoagulant",
  "INSULIN","Insulin",
  "FUROSEMIDE","Diuretic",
  "LISINOPRIL","ACE Inhibitor",
  "ATORVASTATIN","Statin"
)

assign_class <- function(drug_name){

  match <- med_class_map$med_class[
    str_detect(drug_name, med_class_map$pattern)
  ]

  if(length(match) > 0) match[1] else "Other"

}

prescriptions_clean <- prescriptions_clean |>
  mutate(med_class = map_chr(drug_std, assign_class))

prescriptions_clean |>
  count(med_class, sort = TRUE) |>
  gt()

```


6. Build Admission Medication Flags

```{r}
admission_med_flags <- prescriptions_clean |>
  filter(med_class != "Other") |>
  distinct(hadm_id, med_class) |>
  mutate(received = 1L) |>
  pivot_wider(
    names_from = med_class,
    values_from = received,
    values_fill = 0L
  ) |>
  clean_names()

cat("Admissions with classified meds:", nrow(admission_med_flags))

```


7. Final Analytic Dataset

```{r}
analysis_df <- admissions_mortality |>
  inner_join(admission_med_flags, by = "hadm_id") |>
  mutate(
    age_at_admission = anchor_age,
    age_group = cut(
      age_at_admission,
      breaks = c(0,40,55,65,75,Inf),
      labels = c("<40","40–54","55–64","65–74","75+"),
      right = FALSE
    )
  ) |>
  select(
    subject_id,
    hadm_id,
    admittime,
    dischtime,
    los_days,
    gender,
    age_at_admission,
    age_group,
    race,
    mortality_30day,
    days_to_death,
    everything()
  )

cat("Final analytic dataset:", nrow(analysis_df), "rows")

write_csv(analysis_df,"../../data/processed-data/analysis_dataset.csv"
)

```