---
title: "MIMIC-IV Data Processing Pipeline"
author: "Riley Herber and Rebecca BAsta"
date: "2026-02-26"
output: html_document
---

This document covers all data loading, cleaning, and wrangling or the project examining the relationship between medication use and 30-day mortality in the MIMIC-IV database.

**Primary outcome:** 30-day mortality (binary: died within 30 days of admission)  
**Primary predictor:** Medication use (drug name / medication class)  
**Covariates:** Age at admission, gender  


```{r}
library(tidyverse)   # data wrangling & ggplot2
library(lubridate)   # date handling
library(janitor)     # clean_names(), tabyl()
library(skimr)       # skim() for quick summaries
library(gt)          # nice tables
library(here)        # relative file paths

theme_set(theme_minimal(base_size = 13))
```

# 1. Data Loading

The tables used are:

| Table | Description |
|---|---|
| `patients.csv` | Demographics (gender, anchor_age, dod) |
| `admissions.csv` | Admission/discharge times, insurance, ethnicity |
| `prescriptions.csv` | Medication orders per hospital stay |


```{r}
patients      <- read_csv("data/raw-data/patients.csv", show_col_types = FALSE)
admissions    <- read_csv("data/raw-data/admissions.csv", show_col_types = FALSE)
prescriptions <- read_csv("data/raw-data/prescriptions.csv", show_col_types = FALSE)

# work with a sample for speed
# Comment these lines out when ready to run on the full dataset
set.seed(123)
sampled_ids   <- admissions |> slice_sample(n = 10000) |> pull(hadm_id)
admissions    <- admissions    |> filter(hadm_id %in% sampled_ids)
prescriptions <- prescriptions |> filter(hadm_id %in% sampled_ids)
patients      <- patients      |> filter(subject_id %in% admissions$subject_id)

# Clean column names to snake_case
patients      <- clean_names(patients)
admissions    <- clean_names(admissions)
prescriptions <- clean_names(prescriptions)

# Quick dimension check
cat("patients:     ", nrow(patients),      "rows,", ncol(patients),      "cols\n")
cat("admissions:   ", nrow(admissions),    "rows,", ncol(admissions),    "cols\n")
cat("prescriptions:", nrow(prescriptions), "rows,", ncol(prescriptions), "cols\n")
```

---

# 2. Data Cleaning

## 2.1 Patients Table

```{r}
patients_clean <- patients |>
  select(subject_id, gender, anchor_age, anchor_year, dod) |>
  # Ensure dod is parsed as a date
  mutate(dod = as_date(dod))

# Check for missing values
skim(patients_clean)
```

## 2.2 Admissions Table

```{r}
admissions_clean <- admissions |>
  select(subject_id, hadm_id, admittime, dischtime, admission_type,
         admission_location, discharge_location, insurance, race) |>
  mutate(
    admittime  = ymd_hms(admittime),
    dischtime  = ymd_hms(dischtime),
    # Length of stay in days
    los_days   = as.numeric(difftime(dischtime, admittime, units = "days"))
  ) |>
  # Remove admissions with negative or implausible LOS (data entry errors)
  filter(los_days >= 0, los_days < 365)

cat("Admissions after LOS filter:", nrow(admissions_clean), "\n")
```

## 2.3 Compute 30-Day Mortality

30-day mortality is defined as death (recorded in `patients.dod`) occurring within 30 days of the admission date.

```{r}
# Join patient death date onto admissions
admissions_mortality <- admissions_clean |>
  left_join(patients_clean |> select(subject_id, dod, gender, anchor_age),
            by = "subject_id") |>
  mutate(
    # Days from admission to death (NA if patient did not die)
    days_to_death   = as.numeric(difftime(dod, admittime, units = "days")),
    # 30-day mortality flag
    mortality_30day = case_when(
      !is.na(days_to_death) & days_to_death >= 0 & days_to_death <= 30 ~ 1L,
      TRUE ~ 0L
    )
  )

# Sanity check
admissions_mortality |>
  count(mortality_30day) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  gt() |>
  cols_label(mortality_30day = "30-Day Mortality", n = "Count", pct = "Percent (%)")
```

## 2.4 Prescriptions Table

The prescriptions table contains one row per medication order. Key cleaning steps: standardise drug names, remove obvious duplicates, drop rows with missing drug names.

```{r}
prescriptions_clean <- prescriptions |>
  select(subject_id, hadm_id, drug, drug_type, starttime, stoptime,
         route, dose_val_rx, dose_unit_rx) |>
  # Drop rows with no drug name
  filter(!is.na(drug), drug != "") |>
  mutate(
    # Standardise drug names: uppercase, strip extra whitespace
    drug_std = str_squish(str_to_upper(drug)),
    # Parse dates
    starttime = ymd_hms(starttime),
    stoptime  = ymd_hms(stoptime)
  )

cat("Prescriptions after cleaning:", nrow(prescriptions_clean), "\n")
cat("Unique drugs:", n_distinct(prescriptions_clean$drug_std), "\n")
```

### Assign Broad Medication Classes

To enable group-level comparisons, we assign each drug to a broad class using keyword matching. We can always append or remove categories as we see fit.

Note: This next cell usually takes a few minutes to run.
```{r}
# Keyword-to-class mapping
med_class_map <- tribble(
  ~pattern,               ~med_class,
  "ASPIRIN",              "Antiplatelet",
  "CLOPIDOGREL",          "Antiplatelet",
  "HEPARIN",              "Anticoagulant",
  "WARFARIN",             "Anticoagulant",
  "ENOXAPARIN",           "Anticoagulant",
  "METOPROLOL",           "Beta-Blocker",
  "CARVEDILOL",           "Beta-Blocker",
  "ATENOLOL",             "Beta-Blocker",
  "LISINOPRIL",           "ACE Inhibitor",
  "ENALAPRIL",            "ACE Inhibitor",
  "CAPTOPRIL",            "ACE Inhibitor",
  "AMLODIPINE",           "Calcium Channel Blocker",
  "DILTIAZEM",            "Calcium Channel Blocker",
  "FUROSEMIDE",           "Diuretic",
  "SPIRONOLACTONE",       "Diuretic",
  "VANCOMYCIN",           "Antibiotic",
  "PIPERACILLIN",         "Antibiotic",
  "CEFAZOLIN",            "Antibiotic",
  "METRONIDAZOLE",        "Antibiotic",
  "CIPROFLOXACIN",        "Antibiotic",
  "INSULIN",              "Insulin",
  "MORPHINE",             "Opioid",
  "FENTANYL",             "Opioid",
  "HYDROMORPHONE",        "Opioid",
  "OXYCODONE",            "Opioid",
  "NOREPINEPHRINE",       "Vasopressor",
  "VASOPRESSIN",          "Vasopressor",
  "DOPAMINE",             "Vasopressor",
  "EPINEPHRINE",          "Vasopressor",
  "PROPOFOL",             "Sedative",
  "MIDAZOLAM",            "Sedative",
  "LORAZEPAM",            "Sedative",
  "DEXAMETHASONE",        "Corticosteroid",
  "PREDNISONE",           "Corticosteroid",
  "METHYLPREDNISOLONE",   "Corticosteroid",
  "ATORVASTATIN",         "Statin",
  "SIMVASTATIN",          "Statin",
  "ROSUVASTATIN",         "Statin"
)

# Assign class by checking if drug name contains any keyword
assign_class <- function(drug_name) {
  for (i in seq_len(nrow(med_class_map))) {
    if (str_detect(drug_name, fixed(med_class_map$pattern[i]))) {
      return(med_class_map$med_class[i])
    }
  }
  return("Other")
}

prescriptions_clean <- prescriptions_clean |>
  mutate(med_class = map_chr(drug_std, assign_class))

# Class distribution
prescriptions_clean |>
  count(med_class, sort = TRUE) |>
  filter(med_class != "Other") |>
  gt() |>
  cols_label(med_class = "Medication Class", n = "Prescription Count")
```


# 3. Data Wrangling: Build Analysis Dataset

## 3.1 Admission-Level Medication Summary

For each admission, we create binary flags indicating whether the patient
received each medication class during their stay.

```{r}
#| label: admission-meds

# One row per admission × med_class
admission_med_flags <- prescriptions_clean |>
  filter(med_class != "Other") |>
  distinct(hadm_id, med_class) |>
  mutate(received = 1L) |>
  pivot_wider(names_from = med_class, values_from = received,
              values_fill = 0L) |>
  clean_names()

cat("Admissions with at least one classified medication:", nrow(admission_med_flags), "\n")
```

## 3.2 Final Analytic Dataset

This section takes the cleaned mortality data and the cleaned medication data, merges them into one table, trims it down to only what's needed, and saves it. Everything after this chunk is just analysis on that one clean file.
```{r}

analysis_df <- admissions_mortality |>
  inner_join(admission_med_flags, by = "hadm_id") |>
  # Compute age at admission (anchor_age is age at anchor_year)
  mutate(
    age_at_admission = anchor_age,
    age_group = cut(age_at_admission,
                    breaks = c(0, 40, 55, 65, 75, Inf),
                    labels = c("<40", "40–54", "55–64", "65–74", "75+"),
                    right  = FALSE)
  ) |>
  # Keep only relevant columns
  select(subject_id, hadm_id, admittime, dischtime, los_days,
         gender, age_at_admission, age_group, race,
         mortality_30day, days_to_death,
         any_of(tolower(unique(med_class_map$med_class)) |> str_replace_all(" ", "_")))

cat("Final analytic dataset:", nrow(analysis_df), "admissions,",
    ncol(analysis_df), "columns\n")


# Save cleaned dataset
write_csv(analysis_df, "data/processed-data/analysis_dataset.csv")
```
