# Further Analysis and Models
Table 1: Unadjusted odds ratios (95% CI) from single-predictor logistic regression models for 30-day mortality.
```{r}
library(here)
library(dplyr)
library(broom)
library(knitr)

# Load and prepare the analytic dataset
admissions_clean <- read.csv(here("data", "processed-data", "analysis_dataset.csv")) |>
  rename(
    age               = age_at_admission,
    med_anticoagulant = anticoagulant,
    med_statin        = statin,
    med_insulin       = insulin,
    med_antiplatelet  = antiplatelet,
    med_diuretic      = diuretic,
    med_ace_inhibitor = ace_inhibitor
  ) |>
  mutate(
    n_med_classes = med_anticoagulant + med_statin + med_insulin +
                    med_antiplatelet + med_diuretic + med_ace_inhibitor
  )

# Define predictors to loop over
predictors <- c(
  "med_anticoagulant", "med_statin", "med_insulin",
  "med_antiplatelet", "med_diuretic", "med_ace_inhibitor",
  "age", "gender", "n_med_classes"
)

# Fit one logistic regression per predictor and extract tidy results
basic_results <- lapply(predictors, function(pred) {
  formula <- as.formula(paste("mortality_30day ~", pred))
  fit <- glm(formula, data = admissions_clean, family = binomial(link = "logit"))
  tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
    mutate(predictor = pred)
}) |>
  bind_rows() |>
  select(Predictor = predictor, Term = term,
         OR = estimate, `95% CI Lower` = conf.low,
         `95% CI Upper` = conf.high, `p-value` = p.value) |>
  mutate(across(c(OR, `95% CI Lower`, `95% CI Upper`), \(x) round(x, 3)),
         `p-value` = signif(`p-value`, 3))

# Save and display
saveRDS(basic_results, here("results", "tables", "basic_logreg_results.rds"))
kable(basic_results)
```

Figure 5: Forest plot of unadjusted odds ratios (95% CI) for 30-day mortality by predictor. ORs greater than 1 indicate higher odds of mortality; ORs less than 1 indicate lower odds. The dashed vertical line marks the null (OR = 1).


```{r}

library(here)
library(dplyr)
library(broom)
library(ggplot2)

# Load and prepare the analytic dataset
admissions_clean <- read.csv(here("data", "processed-data", "analysis_dataset.csv")) |>
  rename(
    age               = age_at_admission,
    med_anticoagulant = anticoagulant,
    med_statin        = statin,
    med_insulin       = insulin,
    med_antiplatelet  = antiplatelet,
    med_diuretic      = diuretic,
    med_ace_inhibitor = ace_inhibitor
  ) |>
  mutate(
    n_med_classes = med_anticoagulant + med_statin + med_insulin +
                    med_antiplatelet + med_diuretic + med_ace_inhibitor
  )

predictors <- c(
  "med_anticoagulant", "med_statin", "med_insulin",
  "med_antiplatelet", "med_diuretic", "med_ace_inhibitor",
  "age", "gender", "n_med_classes"
)

basic_results <- lapply(predictors, function(pred) {
  formula <- as.formula(paste("mortality_30day ~", pred))
  fit <- glm(formula, data = admissions_clean, family = binomial(link = "logit"))
  tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
    mutate(predictor = pred)
}) |>
  bind_rows() |>
  select(Predictor = predictor, Term = term,
         OR = estimate, `95% CI Lower` = conf.low,
         `95% CI Upper` = conf.high, `p-value` = p.value) |>
  mutate(across(c(OR, `95% CI Lower`, `95% CI Upper`), \(x) round(x, 3)),
         `p-value` = signif(`p-value`, 3))

basic_results |>
  ggplot(aes(x = OR, y = reorder(Term, OR))) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbarh(aes(xmin = `95% CI Lower`, xmax = `95% CI Upper`),
                 height = 0.2, color = "steelblue") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40") +
  scale_x_log10() +
  labs(
    x = "Odds Ratio (log scale)",
    y = NULL,
    title = "Unadjusted Odds Ratios for 30-Day Mortality"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())

ggsave(here("results", "figures", "basic_logreg_forest.png"),
       width = 7, height = 5, dpi = 300)
```

Figure 6: Receiver operating characteristic (ROC) curve for the multivariable logistic regression model predicting 30-day mortality. The AUC reflects the model’s ability to discriminate between survivors and non-survivors.

```{r}
library(pROC)
library(ggplot2)
mv_model <- glm(mortality_30day ~ med_anticoagulant + med_statin + med_insulin +
                 med_antiplatelet + med_diuretic + med_ace_inhibitor + age + gender,
                data = admissions_clean, family = binomial(link = "logit"))

# Generate predicted probabilities (mv_model and admissions_clean carried over from tbl-multivariable)
admissions_clean$pred_prob <- predict(mv_model, type = "response")

# Compute ROC
roc_obj <- roc(admissions_clean$mortality_30day, admissions_clean$pred_prob,
               quiet = TRUE)
auc_val <- round(auc(roc_obj), 3)

# Plot
roc_df <- data.frame(
  specificity = rev(roc_obj$specificities),
  sensitivity = rev(roc_obj$sensitivities)
)

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  annotate("text", x = 0.7, y = 0.15,
           label = paste0("AUC = ", auc_val), size = 4.5) +
  labs(
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)",
    title = "ROC Curve: Multivariable Logistic Regression"
  ) +
  theme_minimal(base_size = 12)

ggsave(here("results", "figures", "roc_curve.png"),
       width = 6, height = 5, dpi = 300)
```

# ML models, cross-validation metrics, and figures

Outputs:
   results/figures/ml_roc_pr_curves.png
   results/figures/ml_model_comparison.png
   results/figures/ml_feature_importance.png
   results/tables/ml_model_comparison.csv


```{r}
# ML pipeline: Medication Regimens & 30-day Mortality
library(tidyverse)
library(tidymodels)
library(ranger)
library(xgboost)
library(rpart)
library(vip)
library(patchwork)
```

```{r}
# Paths
DATA_PATH <- "data/processed-data/analysis_dataset.csv"
FIG_DIR   <- "results/figures"
TBL_DIR   <- "results/tables"
```

```{r}
# Load & preprocess data
df <- read_csv(DATA_PATH, show_col_types = FALSE) |>
  mutate(
    gender_bin = as.integer(gender == "M"),
    polypharmacy = ace_inhibitor + statin + anticoagulant +
      diuretic + antiplatelet + insulin,
    mortality_30day = factor(mortality_30day,
                             levels = c(0, 1),
                             labels = c("survived", "died"))
  ) |>
  select(mortality_30day, age_at_admission, gender_bin,
         ace_inhibitor, statin, anticoagulant,
         diuretic, antiplatelet, insulin, polypharmacy)

cat("Dataset:", nrow(df), "x", ncol(df), "\n")
print(table(df$mortality_30day))
```

```{r}
# CV setup
set.seed(42)
folds <- vfold_cv(df, v = 5, strata = mortality_30day)
metrics <- metric_set(roc_auc, average_precision, brier_class)
```
```{r}
# Recipe
rec_full <- recipe(mortality_30day ~ ., data = df) |>
  step_normalize(age_at_admission, polypharmacy)

rec_age <- recipe(mortality_30day ~ age_at_admission, data = df)

rec_meds <- recipe(
  mortality_30day ~ ace_inhibitor + statin + anticoagulant +
    diuretic + antiplatelet + insulin + polypharmacy,
  data = df
)
```

```{r}
# Models
pos_weight <- sum(df$mortality_30day == "survived") /
              sum(df$mortality_30day == "died")

lr <- logistic_reg() |> set_engine("glm") |> set_mode("classification")

dt <- decision_tree(tree_depth = 4) |> set_engine("rpart") |> set_mode("classification")

rf <- rand_forest(trees = 500) |>
  set_engine("ranger",
             importance = "impurity",
             class.weights = c("survived" = 1, "died" = 1)) |>
  set_mode("classification")

gx <- boost_tree(trees = 200, tree_depth = 3) |>
  set_engine("xgboost",
             scale_pos_weight = pos_weight) |>
  set_mode("classification")

```

```{r}
wf <- function(recipe, model) {
  workflow() |>
    add_recipe(recipe) |>
    add_model(model)
}
```
```{r}
# Workflow
workflows <- list(
  "Logistic Regression" = wf(rec_full, lr),
  "Decision Tree"       = wf(rec_full, dt),
  "Random Forest"       = wf(rec_full, rf),
  "Gradient Boosting"   = wf(rec_full, gx),
  "Age Only"            = wf(rec_age, lr),
  "Medication Only"     = wf(rec_meds, lr)
)
```
```{r}
# Fir CV (single run)
res_list <- map(names(workflows), function(nm) {
  fit_resamples(
    workflows[[nm]],
    resamples = folds,
    metrics = metrics,
    control = control_resamples(save_pred = TRUE)
  )
})
names(res_list) <- names(workflows)
```

```{r}
# Metric Table
summary_tbl <- map_dfr(res_list, collect_metrics, .id = "Model") |>
  select(Model, .metric, mean) |>
  pivot_wider(names_from = .metric, values_from = mean) |>
  rename(
    `AUC-ROC` = roc_auc,
    `Avg Precision` = average_precision,
    `Brier Score` = brier_class
  ) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

write_csv(summary_tbl, file.path(TBL_DIR, "ml_model_comparison.csv"))
print(summary_tbl)
```

```{r}
# Predictions (OOF)
oof_preds <- map_dfr(res_list, collect_predictions, .id = "model")
```

```{r}
# Calibration plot
cal_df <- oof_preds |>
  mutate(bin = ntile(.pred_died, 10)) |>
  group_by(model, bin) |>
  summarise(
    mean_pred = mean(.pred_died),
    obs_rate = mean(mortality_30day == "died"),
    .groups = "drop"
  )

p_cal <- ggplot(cal_df, aes(mean_pred, obs_rate, color = model)) +
  geom_point() +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Calibration Plot", x = "Predicted", y = "Observed") +
  theme_bw()

ggsave(file.path(FIG_DIR, "ml_calibration.png"), p_cal)
```

```{r}
# ROC + PR curves

roc_df <- oof_preds |>
  group_by(model) |>
  roc_curve(mortality_30day, .pred_died, event_level = "second")

p_roc <- ggplot(roc_df, aes(1 - specificity, sensitivity, color = model)) +
  geom_line() +
  geom_abline(linetype = "dashed") +
  theme_bw() +
  labs(title = "ROC Curves")

pr_df <- oof_preds |>
  group_by(model) |>
  pr_curve(mortality_30day, .pred_died, event_level = "second")

p_pr <- ggplot(pr_df, aes(recall, precision, color = model)) +
  geom_line() +
  theme_bw() +
  labs(title = "Precision-Recall Curves")

fig1 <- p_roc + p_pr

ggsave(file.path(FIG_DIR, "ml_roc_pr_curves.png"), fig1)
```


```{r}
# Feature importance (RF)

rf_fit <- wf(rec_full, rf) |> fit(df)

p_imp <- rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 9) +
  theme_bw() +
  labs(title = "Random Forest Feature Importance")

ggsave(file.path(FIG_DIR, "ml_feature_importance.png"), p_imp)

cat("Workflow complete.\n")
```